---
title: "Hand Movements"
output:
  pdf_document: 
    toc: true 
    toc_depth: 3  
    number_sections: true  
    highlight: tango
  html_document: default
  fig_caption: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


\newpage
```{r echo=FALSE}
library(ggplot2)
library(lattice)
library(readxl)
```


# INTRODUCTION

\

People tend to use their hands while talking, at least sometimes. Some people make small hand gestures as a means to visually convey their message. Other people make overly large gestures as a means to interact with their targeted audience. And still others don’t use much hands movements at all. There are many factors that can categorize people into these three categories like age, topic of discussion, fluency of language, targeted audience, self-confidence, stress level etc. The influence of these factors on the hand movements varies with the type of communication under consideration. Therefore, for the purpose of simplification, this observatory study is targeted only on formal communication, that is a ‘presentation’.  

Communication by itself simply does not imply the verbal aspect. Verbal interaction along with body language, facial expressions and hand gestures together comprise a proper communication. As communication plays a vital role in our competitive environment, we believe, that an analysis of the important aspects of communication will be beneficial to all readers. 

The reason for choosing this topic, centers around our mutual interest in a topic area that is very closely related to the present. An inevitable part of a graduate study is presentations. We investigated the various factors that comes into play during a presentation and found that one of the most important features is the hand movements made by the speaker. We agreed upon the thought that females make more hand movements while speaking. So, inorder to make it more interesting, we decided to analyse this very thought, and to find out whether our opinion is right or wrong.

So here, the pivot question under consideration is whether ‘Female speakers are likely to use more hand movements than Male speakers'. To this end, in this project, we are conducting an observational study on the videos of female and male Tedx speakers giving presentations. 

The following sections provides detailed explanation of the methods used in data collection,the type of variables involved, biases or shortcomings present, the parameters and test statistics used and description of the null and alternate hypothesis. Based on these points, we aim to provide an exploratory analysis of this observational study and showcase the desired results around an acceptable confidence interval.
\



## Data Collection


### Sampling stategy:

In this project, the population inferred to, is the population of Tedx speakers in the United States who are getting videotaped. The sampling strategy used here is random sampling. We collected around 50 videos of male and female speakers each. Our collection methodology was implemented by randomly choosing 100 videos of Tedx speakers and assigning a number to each of the selected videos. We have used random sampling to pick 50 videos. In total, we have picked 25 videos of male TedX speakers and 25 videos of female TedX speakers in the United States.

### Uniformity : 

As we are considering the difference in mean number of hand movements between men and women while giving presentations, it is crucial to measure the hand movements for all the videos for the same duration. Therefore, we have fixed the time limit as three minutes for all the videos. For our ur observational study, data points are taken by noting down the number of hand movements of both male and female Tedx speakers during a period of three minutes of their presentations. We have avoided considering the first 1 minute of all the videos, as the speakers may take a minute to get accustomed to the staging and target audience.  


### Shortcomings in Collection Methods:

The following are the shortcomings in our collection methods:

1. As the population inferred in our study is very large, it is not possible to represent the actual stratification in the sample that we have considered.


2. We have not taken into consideration the factors like culture and other external uncontrolled factors like room temperature which can also play a role in the number of hand movements while communication.


3. The presence of two confounding variables namely 'age' and 'topic' in our study.



\


\vspace{1cm}
# EXPLORATORY DATA ANALYSIS

```{r LoadPackages, include=FALSE}

library(lattice)
library(ggplot2)
library(readxl)

```
```{r include=FALSE}

Dataset_Midterm <- read_excel("Dataset_Midterm.xlsx")
dataset <- split(Dataset_Midterm$`Total No. of Hand Movements`,Dataset_Midterm$Gender)

summary(Dataset_Midterm)

barplot(dataset$Male)
barplot(dataset$Female)

histogram(~`Total No. of Hand Movements`|`Gender`, data=Dataset_Midterm, type="count")


`For 25 Male TedX Speakers` <- dataset$Male
`For 25 female TedX Speakers` <- dataset$Female

data <- data.frame(
  type=c(rep("Male",25),rep("Female",25)),
  value=c(dataset$Male[1:25]
          ,dataset$Female)
)

test <- cbind(`For 25 Male TedX Speakers`,
              `For 25 female TedX Speakers`)

```

As we want to study "total no. of hand movements" for both Gender TedX speakers. Let us look at their properties w.r.t to "total no. of hand movements.  

## for Females:
`````{r}
# summary for Females
summary(Dataset_Midterm$`Total No. of Hand Movements`[Dataset_Midterm$Gender == "Female"])
```  
with standard deviation (sd)
```{r}
# standard deviation for Females
sd_Female <- sd(Dataset_Midterm$`Total No. of Hand Movements`[Dataset_Midterm$Gender == "Female"])
sd_Female
```
## for Males:
```{r}
# summary for Males
summary(Dataset_Midterm$`Total No. of Hand Movements`[Dataset_Midterm$Gender == "Male"])
```
with standard deviation (sd)  
```{r}
# Standard Deviation for Males
sd_Male <- sd(Dataset_Midterm$`Total No. of Hand Movements`[Dataset_Midterm$Gender == "Male"])
sd_Male
```

The table shows descriptive statistics for variables. The mean of "total no. of hand movements" for  Female TedX Speakers is higher (86.84) with standard deviation  29.28834 than mean of "total no. of hand movements" for  Male TedX Speakers (48.00) with standard deviation of 19.59592 in our chosen sample.

The observed sample distribution of total number of hand movements for male TedX speakers and female TedX speakers is as follows  

```{r}
# creating Data frame for barplot
df = rbind(Dataset_Midterm$`Total No. of Hand Movements`[Dataset_Midterm$Gender== "Male"]
           ,Dataset_Midterm$`Total No. of Hand Movements`[Dataset_Midterm$Gender== "Female"])

# barplot for Males and Females w.r.t total Hand movements
barplot(df,ylab="Total No. of Hand Movements",ylim=c(0,300)
        ,col=c("blue","green")
        ,xlab="25 Males and Females of sample")
legend("top"
       ,legend=c("Total No. of Hand Movement for Female"
                 ,"Total No. of Hand Movement for Male")
       ,fill=c("green","blue"))
```
  
The above visualistaion shows distribution of observations of "Total No. of Hand Movements" for Male and Female TedX speakers each 25 samples. It suggests that Females have used lot more hand movements than male. 
  
Let us visualize the boxplots for our sample and see if there are any outliers for Male TedX speakers and Female Tedx Speakers.  

```{r}
# boxplot for Male and Female Tedx speakers
boxplot(test, ylab="Total No. of Hand Movements")
```
  
Here we see that the Inter Quantile Range (IQR) for male TedX speakers is smaller than female Tedx speakers, we also have one outlier for male TedX speakers Sample. This boxplot shows that most of the values of “total no. of hand movements” for male TedX speakers are concentrated around 50. The boxplot for female Tedx speakers is higher than Male TedX speakers which shows the values of female Tedx speakers are in higher and wider ranges than male TedX speakers. 

\vspace{1cm}

# STATISTICAL DATA ANALYSIS:

Now we have to make inference to population (Male and Female TedX speakers in United States).  

## Does Female TedX Speakers tends to use more hand gestures than Male TedX Speakers in United States?

### Selection of statistical test

From our question, the main objective is to compare total number of hand movements between males and females. As the comparison is made between two samples of the two different populations. we dont have any information about Population parameters. so, the test that needs to be selected is Two Sample (Male and Female) t-test.


### Population parameter
As we compare "Total No. of Hand Movements" from male TedX speakers and female TedX speakers from sample, we want to know if female TedX speakers uses more hand movements than men or not. For that we need to consider mean of "Total No. of Hand Movements" by female TedX speakers and "Total No. of Hand Movements" by male TedX speakers. So our population parameter is going to be population mean $\mu$.  
  
### Conditions
Before moving ahead with two sample t-test we first need to verify if our sample distribution is following all necessary conditions for t-test.
  
1.  $\textbf{Question of interest has to do with the difference of means between two populations}$  
In our case we have to compare means from female TedX speakers and male TedX speakers. So yes, we will be taking difference of means betwen two populations (male TedX speakers and female TedX speakers).  
  
2.  $\textbf{Two independent samples from two populations}$  
yes. since we are considering male and female TedX speakers as two different populations.  
  
3.  $\textbf{The population data must be normally distributed}$  

\vspace{1cm}
```{r fig.height=3,fig.width=4}
# histogram for Total Hand Movements for Male and Female
par(mfrow=c(2,2))
Dataset_Midterm <- read_excel("Dataset_Midterm.xlsx")
histogram(~`Total No. of Hand Movements`|Gender,Dataset_Midterm, type="count")

# density plot for Total Hand Movements for Male and Female
densityplot(~`Total No. of Hand Movements`|Gender,Dataset_Midterm, type="count")
```
  
\vspace{1cm}
The above distributions tells us that most of the data lies around mean, which makes the distribution appear normal. In order to verify it lets plot qqnorm.  
  
\vspace{1cm}  
```{r fig.height=5}
# qq norm plots for Male and Females
par(mfrow=c(2,2))
qqnorm(Dataset_Midterm[which(Dataset_Midterm$Gender == "Male")
                       ,"Total No. of Hand Movements"]$`Total No. of Hand Movements`
       ,main="For Male tedX speakers")
qqline(Dataset_Midterm[which(Dataset_Midterm$Gender == "Male")
                       ,"Total No. of Hand Movements"]$`Total No. of Hand Movements`)

qqnorm(Dataset_Midterm[which(Dataset_Midterm$Gender == "Female")
                       ,"Total No. of Hand Movements"]$`Total No. of Hand Movements`
       , main="For Female tedX speakers")
qqline(Dataset_Midterm[which(Dataset_Midterm$Gender == "Female")
                       ,"Total No. of Hand Movements"]$`Total No. of Hand Movements`)
```

\vspace{1cm}  
As per qqnorm there are some points which does not fall exactly on $y=x$ line, but the set for both males and females could be used for good approximation for normal distribution.  

# Analysis  

## Parameters  
We are interested to make inference  population mean of total number of hand Movements for TedX speakers, both men and women say $\mu_{w}$ and  $\mu_{m}$ we are interested in $\mu_{w} - \mu_{m}$  

## Hypothesis  
$H_{o}:\mu_{w}=\mu_{m}\space or \space \mu_{w}-\mu_{m} =0$    
The true population mean for "total no. of hand movements for female TedX speakers" is equal to "total no. of hand movements  for male TedX speakers".  

$H_{a}:\mu_{w} > \mu_{m}$ or $\mu_{w}-\mu_{m} > 0$  
The true population mean for "total no. of hand movements for female TedX speakers" is greater than "total no. of hand movements  for male Tedx speakers".

## Sample Statistic  
$$ \bar{x_{w}}- \bar{{x_{m}}}$$  
As we are interested in the statistics describing the variance between the two samples, we need to take into account the standard errors of both samples.  
$$SE_p = \sqrt{\frac{{s_m}^2}{n_m}+\frac{{s_w}^2}{n_w}}$$
and, $n_m = n_w=n$  

The t-test statistic being calculated as,  
$$t_{min(n_w-1,n_m-1)STAT} = \frac{\bar{x_{w}}- \bar{{x_{m}}}}{SE_p}$$
 
```{r}
# Calculating T-stat
stand_error <- sqrt((29.28834)**2/25 +(19.59592)**2/25)
t_stat <- (86.84 - 48)/stand_error
```
$$t_{min(n_w-1,n_m-1)STAT} = `r t_stat`$$   
```{r}
# p-value
one_sided_p_val <-  pt(t_stat,df = min(25,25)-1,lower.tail=FALSE)
```
  
## test-results  
In this case we have considered $\alpha = 0.05$ and will build confidence interval on 95%. since this is one tailed-test we will only consider the right tail or upper tail of the distribution (one sided lower confidence Interval).

Using Student t distribution with degrees of freedom as 24 and $t_{STAT}=`r t_stat`$   we got our p value as $`r one_sided_p_val`$.  

## Confidence Interval  
one-sided lower confidence interval of the difference ($\mu_w$ - $\mu_m$).  
  
$\bar{x_{w}}- \bar{{x_{m}}} + t_{min(n_w-1,n_m-1),\alpha} . SE_{p} < \mu_{m} - \mu_{w} < \infty$    
$26.81916 < \mu_{w}-\mu_{m} < \infty$  

```{r fig.height=5,fig.width=5}
# Confidence Interval for  lower bound
lower<-(86.64-48)+(qt(0.05,25-1)*stand_error)

# plotting Confidence interval and t value on t-distribution.
pts = seq(-10, 10, .005)
plot(pts,dt(pts,df=24),col='red',type='l', main="t distribution")
abline(v=t_stat)
abline(v=qt(0.95,24),col="blue")
text(7, 0.1, "t_stat:5.510896",srt=0.1,pos=3)
text(qt(0.95,24),0.3, "CI lower Bound:t_crit=1.710882",srt=0.1,pos=3)
```
   
The above distribution denotes student t-distribution. The blue line denotes the start of confidence interval and black line denotes the t-value in the distribution. It is clear from the curve that unbounded confidence interval. to the right hand side. since, there is no other blue line it implies that confidence interval extends till infinity.


## Interpretation  
There is some evidence (P-value=`r one_sided_p_val`) that the true population mean of "Total Number of Hand Movements" for Female TedX speakers is greater than Male TedX speakers.  We reject the null hypothesis that there is no difference in the mean of "Total Number of Hand Movements" between the male TedX speakers and female TedX speakers at the $\alpha =0.05$ level. With 95% confidence, the true difference in the mean "total no. of hand movements" between male TedX speakers and female TedX speakers is between $(`r lower`,\infty)$ or the true mean difference in "total No. of Hand Movements" between Female TedX speakers and Male TedX speakers is  `r lower` or above as shown in the graph. 
The null hypothesized difference between the mean of "total no. of hand movements" is zero and zero is not in
the 95% confidence interval which is consistent with the rejection of the null hypothesis and the values of the confidence interval suggest that on average female TedX speakers use more hand movements than male TedX speakers in United States.

\vspace{3cm}
# DISCUSSION:

This observational study was able to provide insights into the relation between gender and hand movements in formal communication.The results suggests, though not in a definitive sense, that there exists a correlation between the gender of the speaker and number of hand movements per minute. This study was able to establish the statement that, the mean number hand movements made by female tedx speakers is greater than the mean number of hand movements made by male tedx speakers. 

However, the limitations of this study includes the fact that, a cause-effect relationship cannot be proven. For example, do females tend to be more interactive than men, or are women more likely to be more proactive than men in genearl? Our analysis does not answer that. Another limitation of this study, is the lack of availability of a more advanced technology to count the exact number of hand movements made by the speakers.We have performed a manual method of counting the hand movements, which can be prone to errors and thus contributing to bias in the collection methodology.Also, our initial sample had 100 random videos from which 50 videos were selected using R sampling methods. But, when a large sample containing thousands of datapoints is considered,a more accurate sampling strategy need to be implemented rather than the normal R random number generator.And lastly, we can cite as a limitation, the fact that we have not considered the videos which are recorded but not yet published.


This study can be concluded by addressing the question that can arise: what would be the result, if all the participants are put in the same circumstances and  their hand movements are counted?.For example, a more focussed study of both men and women speaking on the same topic.


# APPENDIX

Data source: [TedX videos](https://www.ted.com/talks)
```{r ref.label=knitr::all_labels(), echo = T, eval = F}

```








 

  